Move Schedules: Fast persistence computations in coarse dynamic settings\thanks{\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}

### 2021-11-02 move_schedules.tex
Changed conclusion and updated keywords and AMS subject classifications

% ------  Begin document ------- 
\begin{document} 
\maketitle

@@

% REQUIRED
\begin{keywords}
  Algorithms, 
\end{keywords}

@@

Despite its theoretical efficiency, vineyards is often not the method of choice in practical settings. 
While there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc. See~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 has an implementation of vineyards, but the algorithm was never ported to the 2nd version.} 
The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, whether performing the reductions naively or updating the decomposition dynamically using vineyards, they observed maintaining the decomposition is the most computationally demanding aspect of RIVET \cite{rivet}.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling   \emph{continuous} time families of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, dramatically reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases.

 %TODO fill in section reference above.

@@

\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et. al.~\cite{busaryev2010tracking} to include  coarser  operations---in terms of the number of valid intermediate decomposition states, as compared to vineyards---for dynamic persistence. 
We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case bounds in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d \log d)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistent homology computations is given; 
our move schedules are demonstrated to be an order of magnitude more efficient than existing approaches in applications such as flock simulations and 2D persistence computations with RIVET. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

\begin{algorithm}[h]
	\caption{Scheduling algorithm}\label{alg:schedule}
	\setstretch{1.05}
    \begin{algorithmic}[1]
    	\Require Ordered set of filtrations $\mathcal{K}$, each of size $m$, with bijections $f_{i}: K_i \to K_{i+1}$
    	\Ensure Valid $R = D V$ decompositions are computed for each filtration $K_1, K_2, \dots, K_n$
    	\Procedure{CoarseSchedule}{$\mathcal{K} = ( K_1, K_2, \dots, K_n )$} 
    		\State $\mathcal{S} = \emptyset$
    		\For{$i = 1$\textbf{ to }$n-1$}
    		    \State $(p, q) \gets ([m], \, \mathrm{Im}\, f^\ast_{i})$ 
    			\State $\mathrm{lis}_q \gets \mathrm{LIS}(q)$ \Comment{$O(m \log \log m)$}
    			\State $\mathcal{S} \gets \mathcal{S} \cup \mathrm{GreedySchedule}(p, q, \mathrm{lis}_q)$ \Comment{$O(d\log m)$, see section~\ref{sec:proxy_objective}}
    		\EndFor
	    	\State $(R, V) \gets$ \Call{Reduction}{$D = \partial K_1$}
	    	\For{$(i,j)$\textbf{ in }$\mathcal{S}$}
	    		\If{$i < j$}
	    			\State $(R, V) \gets$ \Call{MoveRight}{i, j}
	    		\Else
	    			\State $(R, V) \gets$ \Call{MoveLeft}{i, j}
	    		\EndIf
	    	\EndFor
    	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\noindent Algorithm~\ref{alg:schedule} is purely illustrative at this point, 
and is meant to serve as a guidepost before discussing its  components in depth. 
Specifically, lines (4-5) are discussed next in section~\ref{sec:schedule_sizes}, and the algorithm for line (6) is given in section~\ref{sec:schedule_cost}. 
Note that the loop starting at line (3) in Algorithm~\ref{alg:schedule} could always be merged with the loop starting at line (8), i.e. the individual schedules $S_i$ may be computed on the fly. Due to the combinatorial aspect of our approach, there is actually no need to have access to---or explicitly store---the entire family of filtrations and the schedules between them. That is, Algorithm~\ref{alg:schedule} can be  modified to be completely online, keeping at most two filtrations and one decomposition in memory at any given time.

@@

	\mathrm{d}_{\mathrm{lcs}}(p,q) = m + n - 2\lvert LCS(p, q)\rvert

@@

F(s_\sigma \circ p, q) = 

@@

	 \mathcal{A}(\sigma) + \lvert i - j \rvert & p^{-1}(\sigma) = i \\

@@

It is not immediately clear how to achieve the latter modification complexity in $O(\log m)$ time, since $\lvert i - j \rvert \leq m$ is potentially larger than $O(\log m)$. %, and 

@@

Since single element modifications, removals, and insertions can all be achieved in $O(\log m)$ expected time with this data structure, we conclude that a greedy solution to equation~\ref{eq:greedy_step} can be computed in $O(d^2 \log m)$ time. 

@@

\section{Conclusion and Future Work}\label{sec:conclusion}
In conclusion, we presented a scheduling algorithm for efficiently updating a decomposition in coarse dynamic settings. Our approach is simple, easy to implement, and fully general: it does not depend on the geometry of underlying space nor the choice of triangulation. Moreover, we supplied efficient algorithms for our scheduling strategy, provided tight bounds where applicable, and demonstrated our algorithms performance with several real world use cases.

There are many possible applications of our work beyond the example application of generating crocker stacks and simulating multidimensional persistence. As discussed in sections~\ref{sec:related_work}, examples include: 
\begin{enumerate}
	\item Computing persistent images for time-varying systems 
	\item Optimization procedures involving persistence diagrams
	\item Detecting homological critical points in time-varying filtrations
\end{enumerate}
Indeed, we envision our approach as potentially useful to any situation where the structure of interest is a parameterized family of filtrations and the goal involves computing the persistent homology for each filtration in the family. Areas of particular interest include time-series analysis and dynamic metric spaces~\cite{kim2020spatiotemporal}.

The simple and combinatorial nature of approach does pose some limitations to its applicability. There may be better bounds or algorithms available if stronger assumptions can be made on how the filtration is changing. If the filtration $K_0$ shares little similarity to the ``target'' filtration $K_1$, then the overhead of reducing the simplices from $K_1 \setminus K_0$ appended to the decomposition derived from $K_0$ may be large enough to motivate simply computing the decomposition at $K_1$ independently, especially if parallel processors are available. This overhead does not exist for the multidimensional persistence use-case: the entire decomposition is needed at each template 2-cell.

@@

Moving forward, our results suggest there are many aspects of computing persistence in dynamic settings yet to be explored. For example, it's not immediately clear whether one could adopt, for example, the twist optimization used in the reduction algorithm to the dynamic setting. Another avenue of interest is the setting where one has knowledge a priori of the time-varying... particular spaces

@@

\appendix 
\section{}\label{sec:appendix}
\subsection*{Vineyard Algorithm}
\begin{algorithm}[!htb]
	\caption{Vineyards: Tranposition Framework}\label{alg:tr}
	\setstretch{1.15}
	\begin{algorithmic}[1]
		\Require Matrices $R, V$ satisfying $R = D V$, $1 \leq i \leq m - 1$
		\Ensure Output $(R, V)$ maintains the decomposition invariants
        \Function{Transpose}{$R$, $V$, $i$}
            \State $\mathrm{pos} \gets $ columns satisfying $\mathrm{col}_R = 0$
            \If {$\mathrm{pos}[i]$ \textbf{and} $\mathrm{pos}[i+1]$}
	            \If{$V[i,i+1] \neq 0$}
            		\State $\mathrm{col}_V(i+1)  \mathrel{+}= \mathrm{col}_V(i)$	
            	\EndIf
            	\If {$\exists \; k, l$ satisfying $\mathrm{low}_R(k) = i, \mathrm{low}_R(l) = i+1$ and $R[i,l] \neq 0$} \Comment{O(m)}
            		\If {$k < l$}
            			\State \Return $(\, R, \, V \,) \gets (\, PRPS_k^l, \, PVP S_k^l \,)$
            		\Else
            			\State \Return $(\, R, \, V \,) \gets  (\, PRPS_l^k, \, PVP S_l^k\,) $
            		\EndIf
            	\EndIf 
            \ElsIf{$!\mathrm{pos}[i]$ \textbf{and} $!\mathrm{pos}[i+1]$}
            	\If{$V[i,i+1] \neq 0$}\Comment{O(m)}
            		\If{$\mathrm{low}_R(i) < \mathrm{low}_R(i+1)$}
            			\State \Return  $(\, R, V \,) \gets (P R S^{i+1}_i P, \, P V S^{i+1}_i  P) $
            		\Else 
            			\State \Return  $(\, R, V \,) \gets (P R S^{i+1}_i P S^{i+1}_i, \, P V S^{i+1}_i P S^{i+1}_i ) $
            		\EndIf 
            	\EndIf
          	\ElsIf{$!\mathrm{pos}[i]$ \textbf{and} $\mathrm{pos}[i+1]$} 
            	\If{$V[i,i+1] \neq 0$}\Comment{O(m)}
            		\State \Return  $(\, R, V \,) \gets (P R S^{i+1}_i P S^{i+1}_i, \, P V S^{i+1}_i P S^{i+1}_i ) $
            	\EndIf
            \ElsIf{$\mathrm{pos}[i]$ \textbf{and} $!\mathrm{pos}[i+1]$}
            	\If{$V[i,i+1] \neq 0$}\Comment{O(m)}
            		\State $\mathrm{col}_V(i+1) \mathrel{+}= \mathrm{col}_V(i)$
            	\EndIf 
            \EndIf
            \State \Return $(\, R , V \,) \gets (\, P R P , P V P \,)$
        \EndFunction
    \end{algorithmic}
\end{algorithm}

>>>>

% ------  Begin document ------- 
% https://github.com/peekxc/dart
\begin{document} 
\maketitle

@@

% REQUIRED
\begin{keywords}
  Computational topology, Persistent homology, Topological data analysis
\end{keywords}

@@

Despite its theoretical efficiency, vineyards is often not the method of choice in practical settings. 
While there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc. See~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 does have an implementation of vineyards, however the algorithm was never ported to version 2. Other major packages, such as GUDHI and PHAT, do not have vineyards implementations.} 
The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, whether performing the reductions naively or updating the decomposition dynamically using vineyards, they observed maintaining the decomposition is the most computationally demanding aspect of RIVET \cite{rivet}.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling   \emph{continuous} time families of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, dramatically reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is available online as an open source \texttt{R} package.\footnote{ Code is available here: \url{https://github.com/peekxc/dart}}

@@

\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et. al.~\cite{busaryev2010tracking} to include  coarser  operations---in terms of the number of valid intermediate decomposition states, as compared to vineyards---for dynamic persistence. 
We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case bounds in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d^2 \log m)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistent homology computations is given; 
our move schedules are demonstrated to be an order of magnitude more efficient than existing approaches in applications such as flock simulations and 2D persistence computations with RIVET. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

\begin{algorithm}[h]
	\caption{Scheduling algorithm}\label{alg:schedule}
	\setstretch{1.05}
    \begin{algorithmic}[1]
    	\Require Ordered set of filtrations $\mathcal{K}$, each of size $m$, with bijections $f_{i}: K_i \to K_{i+1}$
    	\Ensure Valid $R = D V$ decompositions are computed for each filtration $K_1, K_2, \dots, K_n$
    	\Procedure{CoarseSchedule}{$\mathcal{K} = ( K_1, K_2, \dots, K_n )$} 
    		\State $\mathcal{S} = \emptyset$
    		\For{$i = 1$\textbf{ to }$n-1$}
    		    \State $(p, q) \gets ([m], \, \mathrm{Im}\, f^\ast_{i})$ 
    			\State $\mathrm{lis}_q \gets \mathrm{LIS}(q)$ \Comment{$O(m \log \log m)$}
    			\State $\mathcal{S} \gets \mathcal{S} \cup \mathrm{GreedySchedule}(p, q, \mathrm{lis}_q)$ \Comment{$O(d^2\log m)$, see section~\ref{sec:proxy_objective}}
    		\EndFor
	    	\State $(R, V) \gets$ \Call{Reduction}{$D = \partial K_1$}
	    	\For{$(i,j)$\textbf{ in }$\mathcal{S}$}
	    		\If{$i < j$}
	    			\State $(R, V) \gets$ \Call{MoveRight}{i, j}
	    		\Else
	    			\State $(R, V) \gets$ \Call{MoveLeft}{i, j}
	    		\EndIf
	    	\EndFor
    	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\noindent Algorithm~\ref{alg:schedule} is purely illustrative at this point, 
and is meant to serve as a guidepost before discussing its  components in depth. 
Specifically, lines (4-5) are discussed next in section~\ref{sec:schedule_sizes}, and the algorithm for line (6) is given in section~\ref{sec:schedule_cost}. 
Note that the loop starting at line (3) in Algorithm~\ref{alg:schedule} could always be merged with the loop starting at line (8), i.e. the individual schedules $S_i$ may be computed on the fly. Due to the combinatorial aspect of our approach, there is actually no need to have access to---or explicitly store---the entire family of filtrations and the schedules between them. That is, Algorithm~\ref{alg:schedule} can be  modified to be completely online, keeping at most two filtrations and one decomposition in memory at any given time.

@@

	\mathrm{d}_{\mathrm{lcs}}(p,q) = m + n - 2\lvert \mathrm{LCS}(p, q)\rvert

@@

\mathcal{A}(s_\sigma \circ p) = 

@@

	 \mathcal{A}(\sigma) \pm \lvert i - j \rvert & p^{-1}(\sigma) = i \\

@@

It is not immediately clear how to achieve the latter modification complexity in $O(\log m)$ time, since $\lvert i - j \rvert \leq m$ is potentially larger than $O(\log m)$.  

@@

Since single element modifications, removals, and insertions can all be achieved in $O(\log m)$ expected time with such a data structure, we conclude that a greedy solution to equation~\ref{eq:greedy_step} can be computed in $O(d^2 \log m)$ time. 

@@

\section{Conclusion and Future Work}\label{sec:conclusion}
In conclusion, we presented a scheduling algorithm for efficiently updating a decomposition in coarse dynamic settings. Our approach is simple, relatively easy to implement, and fully general: it does not depend on the geometry of underlying space, the choice of triangulation, or the choice of homology dimension. Moreover, we supplied efficient algorithms for our scheduling strategy, provided tight bounds where applicable, and demonstrated our algorithms performance with several real world use cases.

There are many possible applications of our work beyond the example application of generating crocker stacks and simulating multidimensional persistence. As discussed in sections~\ref{sec:related_work}, examples include: 
\begin{enumerate}
	\item Accelerating PH featurization methods for time-varying systems 
	\item Optimization procedures involving persistence diagrams
	\item Detecting homological critical points in time-varying filtrations
\end{enumerate}
Indeed, we envision our approach as potentially useful to any situation where the structure of interest is a parameterized family of filtrations and the goal involves computing the persistent homology for each filtration in the family. Areas of particular interest include time-series analysis and dynamic metric spaces~\cite{kim2020spatiotemporal}.

The simple and combinatorial nature of our approach does pose some limitations to its applicability. For example, there may be better bounds or algorithms available if stronger assumptions can be made on how the filtration is changing with time. Moreover, if the filtration $K_0$ shares little similarity to the ``target'' filtration $K_1$, then the overhead of reducing the simplices from $K_1 \setminus K_0$ appended to the decomposition derived from $K_0$ may be large enough to motivate simply computing the decomposition at $K_1$ independently, especially if parallel processors are available. That is, our approach is primarily useful if each adjacent filtration in the ordered family of filtration is ``close.'' 
Another limitation to the strategy we have here is that we rely completely on the reduction algorithm for appending new simplices to the filtration. If the dynamic filtration is constantly both permanently removing and appending new simplices, then the size of the LIS may very well be on the order $O(m - \sqrt{m})$, rendering the efficiencies gained using our approach as neglible. Note that this overhead does not exist for the multidimensional persistence use-case: the entire decomposition is needed at each template 2-cell.

@@

Moving forward, our results suggest there are many aspects of computing persistence in dynamic settings yet to be explored. For example, it's not immediately clear whether one could adopt, for example, the twist optimization~\cite{chen2011persistent} used in the reduction algorithm to the dynamic setting. Another direction to explore would be the analysis of our approach under the cohomology computation~\cite{de2011dualities}, or the specialization of the move operations to specific types of filtrations such as Rips filtrations. Such adaptations may result in even larger reductions in the number of column operations, as have been observed in practice for the standard reduction algorithm~\cite{bauer2021ripser}.

@@

\appendix 
\section{}\label{sec:appendix}
\subsection*{Vineyard Algorithm}
\begin{algorithm}[!htb]
	\caption{Vineyards: Tranposition Framework}\label{alg:tr}
	\setstretch{1.15}
	\begin{algorithmic}[1]
		\Require Matrices $R, V$ satisfying $R = D V$, $1 \leq i \leq m - 1$
		\Ensure Output $(R, V)$ maintains the decomposition invariants
        \Function{Transpose}{$R$, $V$, $i$}
            \State $\mathrm{pos} \gets $ columns satisfying $\mathrm{col}_R = 0$
            \If {$\mathrm{pos}[i]$ \textbf{and} $\mathrm{pos}[i+1]$}
	            \If{$V[i,i+1] \neq 0$}
            		\State $\mathrm{col}_V(i+1)  \mathrel{+}= \mathrm{col}_V(i)$	
            	\EndIf
            	\If {$\exists \; k, l$ s.t. $\mathrm{low}_R(k) = i, \mathrm{low}_R(l) = i+1$ and $R[i,l] \neq 0$} \Comment{O(m)}
            		\If {$k < l$}
            			\State \Return $(\, R, \, V \,) \gets (\, PRPS_k^l, \, PVP S_k^l \,)$
            		\Else
            			\State \Return $(\, R, \, V \,) \gets  (\, PRPS_l^k, \, PVP S_l^k\,) $
            		\EndIf
            	\EndIf 
            \ElsIf{$!\mathrm{pos}[i]$ \textbf{and} $!\mathrm{pos}[i+1]$}
            	\If{$V[i,i+1] \neq 0$}\Comment{O(m)}
            		\If{$\mathrm{low}_R(i) < \mathrm{low}_R(i+1)$}
            			\State \Return  $(\, R, V \,) \gets (P R S^{i+1}_i P, \, P V S^{i+1}_i  P) $
            		\Else 
            			\State \Return  $(\, R, V \,) \gets (P R S^{i+1}_i P S^{i+1}_i, \, P V S^{i+1}_i P S^{i+1}_i ) $
            		\EndIf 
            	\EndIf
          	\ElsIf{$!\mathrm{pos}[i]$ \textbf{and} $\mathrm{pos}[i+1]$} 
            	\If{$V[i,i+1] \neq 0$}\Comment{O(m)}
            		\State \Return  $(\, R, V \,) \gets (P R S^{i+1}_i P S^{i+1}_i, \, P V S^{i+1}_i P S^{i+1}_i ) $
            	\EndIf
            \ElsIf{$\mathrm{pos}[i]$ \textbf{and} $!\mathrm{pos}[i+1]$}
            	\If{$V[i,i+1] \neq 0$}\Comment{O(m)}
            		\State $\mathrm{col}_V(i+1) \mathrel{+}= \mathrm{col}_V(i)$
            	\EndIf 
            \EndIf
            \State \Return $(\, R , V \,) \gets (\, P R P , P V P \,)$
        \EndFunction
    \end{algorithmic}
\end{algorithm}
